---
title: 'Peer-graded Assignment: Course Project 1'
author: "Shounak Shastri"
date: "18/04/2020"
output:
  html_document: default
  pdf_document: default
  keep_md: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

Here, we will load the dataset and check the summary. 

I had already extracted the data from the zipped file when I downloaded it. 
So I will directly load the data from my working directory in this code chunk.

```{r, echo = TRUE}
setwd("C:/Users/Intel/Documents/Datasets/Activity")
activity <- read.csv("activity.csv")
summary(activity)
```

From the above summary, we can see that the "Steps" column contains NAs. We will
keep them as it is and while calculating the mean and median, we will ignore NA.


## What is the mean total number of steps take per day ?

### We will first make a histogram of the total number of steps taken per day.

After scrolling through the data, we can see that each day is divided into
multiple intervals and the "steps" column shows the number of steps taken in
each interval.

To create a histogram, we will create a new table where we drop the "intervals"
column and display only the total number of steps taken every day.

```{r, echo = TRUE}
totalStepsPerDay <- aggregate(steps ~ date, activity, sum, na.rm=TRUE)
head(totalStepsPerDay)
```

Here, we can see that the new table has dropped the entry for the date
2012-10-01. This is because all the entries in the steps column for this day
were NA.

Now we will plot the histogram

```{r, echo=TRUE}
hist(totalStepsPerDay$steps, main = "Histogram of total steps taken each day",
     xlab = "Total Steps", 
     ylim = c(0, 30))

```

### Calculating and displaying the mean and the median

```{r, echo=TRUE}
cat("The mean is: ", mean(totalStepsPerDay$steps), "\n̥")
cat("The median is: ", median(totalStepsPerDay$steps))
```

## What is the average daily activity pattern?

Here we need to plot the time series plots for average number of steps taken.
But these plots are with respect to the intervals and not dates. So we create a
new table which drops the date column and contains the average number of steps 
for each interval. 

We do this in the same way we made the totalStepsPerDay table above.

```{r, echo=TRUE}
avgStepsPerInterval <- aggregate(steps ~ interval, activity, mean, na.rm=TRUE)
head(avgStepsPerInterval)
```

Now we plot the average steps

```{r, echo=TRUE}

plot(avgStepsPerInterval$interval, avgStepsPerInterval$steps, type = "l", 
     main = "Time series plot for avg steps taken per interval", 
     xlab = "Interval",
     ylab = "avg Steps")

```

Now, we need to find the interval which has the maximum number of average steps

We can see from the above figure that the max number occurs somewhere between 
500 and 1000. This would serve as a good point to check the validity of our
result.

```{r, echo=TRUE}
maxStepsInterval <- avgStepsPerInterval[which.max(avgStepsPerInterval$steps), ]$interval
cat("The interval with the max average steps is: ", maxStepsInterval)
```

We can see the answer is between 500 and 1000 as we discussed above. This means 
our analysis is correct.

## Inputting missing values

We saw from our first code chunk that there are several NA values in the steps
column of our original data. Here we will replace the NA values with some number

### Strategy for replacing the NA values
As the NA values are in the "steps" column, we can take replace them in two ways

1. Make all the NA values as 0.
  This will indicate that the person is sedentary during the whole day. Although
  this seems a resonable option, this shouldn't be considered here as it is 
  impossible to not take a single step in a whole day for a person without any 
  serious disabilities. This option will skew our data to the left and it would 
  be easier if we drop the first day completely.

2. Another option is to replace the values with the average values for that 
  interval. We already have this data in the above code chunk. This might also
  a data which a closer to the actual scenario.

```{r, echo = TRUE}
# First copy the original data in a new table
cleanedActivity <- activity

# Check to see if the steps value is NA and replace it with average value if
# TRUE

for(i in 1:nrow(cleanedActivity)){
    if(is.na(cleanedActivity[i,]$steps)){
        cleanedActivity[i,]$steps <- avgStepsPerInterval[cleanedActivity[i,]$interval == avgStepsPerInterval$interval, ]$steps
    }
}
```

On checking the data, we can see that the NAs in step column have been replaced 
average steps according to the interval.

But, we can also see that some values are floats and some are integers. We need 
a to make all values as integers as it would simplify further analysis.

```{r, echo= TRUE}
cleanedActivity$steps <- as.integer(cleanedActivity$steps)
class(cleanedActivity$steps)
```

Now, our new data set is ready.

### Plotting the histogram for new data
```{r, echo=TRUE}
totalStepsPerDayClean <- aggregate(steps ~ date, cleanedActivity, sum, na.rm=TRUE)

hist(totalStepsPerDayClean$steps, 
     main = "Histogram of total steps taken each day after cleaning data",
     xlab = "Total Steps", 
     ylim = c(0, 40))
```

### Calculating the mean and median of cleaned data

```{r, echo=TRUE}
cat("The mean is: ", mean(totalStepsPerDayClean$steps), "\n̥")
cat("The median is: ", median(totalStepsPerDayClean$steps))
```

Here the difference between our cleaned data and the previous data is 

```{r, echo=TRUE}
cat("difference between means: ", mean(totalStepsPerDayClean$steps) - mean(totalStepsPerDay$steps), "\n")
cat("difference between medians: ", median(totalStepsPerDayClean$steps) - median(totalStepsPerDay$steps), "\n")
```

The negative values indicate that our distribution has shifted to the left after
cleaning the data. This goes according to our intuition as more values
are considered.

## Are there differences in activity patterns on Weekdays and Weekends?

To check this, we first need to check the format of the date column of the 
cleaned data.

```{r, echo=TRUE}
class(cleanedActivity$date)
```

Observe that it is a factor. We need to convert it to date-time data so that we
can use the Weekdays() function.

```{r, echo=TRUE}
cleanedActivity$date <- as.Date(strptime(cleanedActivity$date, format="%Y-%m-%d"))
```

Now, we can use the Weekdays function and convert all the dates to either 
"Weekdays or "Weekends" using a for loop.

```{r, echo=TRUE}
cleanedActivity$day <- weekdays(cleanedActivity$date)
for (i in 1:nrow(cleanedActivity)) {
    if (cleanedActivity[i,]$day == "Saturday" | cleanedActivity[i,]$day == "Sunday") {
        cleanedActivity[i,]$day<-"weekend"
    }
    else{
        cleanedActivity[i,]$day<-"weekday"
    }
}
WeekdayVSWeekend <- aggregate(cleanedActivity$steps ~ cleanedActivity$interval + cleanedActivity$day, cleanedActivity, mean)

head(WeekdayVSWeekend)
```

Give proper names to the columns and conver the steps into integers

```{r, echo=TRUE}
names(WeekdayVSWeekend) <- c("interval", "dayType", "steps")
WeekdayVSWeekend$steps <- as.integer(WeekdayVSWeekend$steps)
head(WeekdayVSWeekend)
```

Now we will plot the graphs

```{r, echo = TRUE}
library(lattice)

xyplot(steps ~ interval | dayType,
       WeekdayVSWeekend, type = "l",
       layout = c(1, 2), 
       xlab = "Interval",
       ylab = "Number of steps")
```